Simulation of an Attic Fire in a Wood Frame Residential Structure - Chicago, IL

### 2015-08-31 Studies/2012_Chicago/Report/firetech_chicago.tex
Chicago: updates to firetech version
Update figure, update firetech version

\input{../../../Bibliography/commoncommands}
\renewcommand{\sb}[1]{_\mathrm{#1}}
\renewcommand{\C}{\mbox{C}}
\renewcommand{\H}{\mbox{H}}
\renewcommand{\O}{\mbox{O}}
\newcommand{\N}{\mbox{N}}

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\chead{}
\renewcommand{\headrulewidth}{0pt}

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2}~story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are:
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.1) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{app_draw} contains dimensioned drawings of the first and second floor of the structure.

@@

\subsection{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 210~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Stair 3 directly asceneds to attic without a doorway or method to isolate the rear porch from the attic (the top stair is the floor of the attic). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

@@

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion for other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing and extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users with the option to use a more complex finite-rate combustion mechanism, there is not sufficient evidence in this case to justify deviating from the default specifications. The product yeilds used in this analysis (Eq.~\ref{eq:wood_comb}) are for well-ventilated burning conditions. The simple chemistry reaction mechanism in FDS assumes constant, user-specified yeilds for carbon monoxide and soot. When a fire becomes underventilated, the yeilds of soot, carbon monoxide, or other product species change due to incomplete combustion. If the analysis of model results requires species concentrations, different species yeilds, a more complex chemical reaction, or finite-rate chemistry may be needed~\cite{FDS_Math_Guide}. The focus of this analysis is to examine the the effect on the flow conditions as a result of a interior door failure. Therefore, errors in species concentrations are tolerated.

To estimate the heat release rate per unit area (HRRPUA) of wood, Babrauskas and Grayson~\cite{babrauskas1990} conducted experiments in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials~\cite{ASTM:E1355}.} to determine the 5-min average of the HRRPUA for several different types of wood over a range of radiant heat fluxes. The results of that study indicate that the HRRPUA for different types of wood is approximately 50~kW/m$^2$. This value may be a conservative estimate for the HRRPUA for wood. As a result, the actual fire may have been larger. However, as discusssed in this study, the fire used in the model produced interior conditions that were hazardous to fully protected firefighers.

@@

\subsection{Materials}
\label{matl}
While the fires were represented as fuel sources with a constant area (Section~\ref{fire}), from a heat transfer perspective, it is important to define the material properties (density, thermal conductivity, and specific heat) to account for heat transfer and energy storage in the ceiling, walls, and floor. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings on the main two floors of the structure, and the material properties of wood~\cite{Incropera:1} were specified on the unfinished portions of the attic and rear porch. Note that the multi-component heat transfer that would typically occur in an exterior wall (gypusum, wood studs, insultation, exterior finish, air gaps) was simplified and respresented by a single material. The following lines define these materials in the FDS input file:

@@

In Fig.~\ref{fig:hrr}, the solid line represents the baseline FDS simulation (this follows the noted door openings from the incident). The dashed line represents an alternative series of events in which the rear door does not get opened and the subsequent breach of the rear porch wall on the left side of the structure does not occur. The dash-dotted line represents the prescribed (or nominal) HRR that was input into FDS. As shown in Fig.~\ref{fig:hrr}, the HRR output from FDS is less than the prescribed HRR because the fire in the structure becomes ventilation limited. The maximum HRR value is slightly less than 9~MW before the fire becomes ventilation limited and begins to decay. The HRR is reduced to 5~MW as the oxygen inside the structure is consumed. There is an insufficient amount of oxygen in the structure to burn all of the fuel, therefore combustion occurs locally at the vents where fresh air is available. Figure~\ref{fig:smv_ext_fire} shows flaming combustion through the failed window on the rear side of the structure and attic breach just after the second fire ignites at 40~s. Figure~\ref{fig:smv_ext_fire} also shows a photograph taken during an attic fire experiment that cofirms this type of fire behavoir. After additional wall breaches occur at 45~s and 55~s (see Table~\ref{tab:vents}), the abrupt change in ventilation causes the HRR to increase to 9~MW before it decreases to approximately 6~MW. The HRR then increases towards a sustainable steady-state value.

Figure~\ref{fig:hrr} indicates that the HRRs for both of the ventilation scenarios in FDS are identical until 130~s (HRR $\approx$ 7~MW). At 130~s, the rear door is opened in the baseline simulation. At 135~s, there is an additional breach on the left side of the structure (Table~\ref{tab:vents}). The open door and breach act as additional pressure releases for the structure. Again, the abrupt change in ventilation causes the HRR to increase from approximately 7~MW to 9.7~MW at 139~s. At the time of the door failure (160~s), the HRR in both cases was approximately 9~MW because of the increased amount of ventilation. While there might be some locally different burning characteristics between the baseline simulation and the alternative simulation, with respect to exposure to firefighters operating within flow path, both cases presented similar high hazard contitions. Therefore, the analysis will focus on the results of the baseline simulation.

\begin{figure}[!ht]
\centering
\includegraphics[width=.5\textwidth]{../Figures/smv_exterior_fire}
\includegraphics[width=.45\textwidth]{../Figures/PA020413}
\caption{Smokeview rendering of exterior combustion occurring on the rear side of the structure (left) and photograph from ongoing attic fire experiments (right).}
\label{fig:smv_ext_fire}
\end{figure}

\subsection{Pressure}
\label{pres}
As the simulated fire grows in the attic and rear porch, the pressure rises. Figure~\ref{fig:pres_159s} shows the calculated pressure conditions in the structure just before the interior door failed in the simulation. On the rear porch, the pressure rise ranges from a 4~Pa ($5.8 \times 10^{-4}$~psi) over-pressure at a height of 1~m to greater than 10~Pa ($1.5 \times 10^{-3}$~psi) at the ceiling from bouyancy due to temperature differences. Gases at an elevated pressure will always move towards a region of lower pressure.The door serves as a boundary between the high and low pressure regions. Once the door fails open, the high pressure gases flow towards the low pressure regions in the house, hallway, kitchen, and front stairwell. The velocity vectors shown in Fig.~\ref{fig:velo_175s} (15~s after the interior door fails) indicate the direction of flow within the structure.

>>>>

\usepackage{pdfsync,cite,hyperref}
\usepackage{tikz,pgfplots}
\usepackage{rotating,lscape}
\usepackage{listings}
\usepackage{multirow,tabularx,ragged2e,booktabs,caption,graphicx,longtable}
\usepackage{mathptmx,xfrac,amsmath,siunitx}

\pgfplotsset{
  colormap={blackwhite}{[5pt]
    rgb255(0pt)=(0,0,255);
    rgb255(100pt)=(0,255,255);
    rgb255(200pt)=(0,255,0);
    rgb255(300pt)=(255,255,0);
    rgb255(400pt)=(255,0,0)
  },
} % sets colors for colorbar

\renewcommand{\sb}[1]{_\mathrm{#1}}
\newcommand{\C}{\mbox{C}}
\renewcommand{\H}{\mbox{H}}
\renewcommand{\O}{\mbox{O}}
\newcommand{\N}{\mbox{N}}
\newcommand{\dx}{\delta x}
\newcommand{\dQ}{\dot{Q}}

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2}~story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are:
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.1) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Fully dimensioned drawings of the first and second floor of the structure can be found in \cite{Weinschenk:Chicago}.

@@

\subsection{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 210~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Stair 3 directly ascends to attic without a doorway or method to isolate the rear porch from the attic (the top stair is the floor of the attic). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

@@

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion for other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing and extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users with the option to use a more complex finite-rate combustion mechanism, there is not sufficient evidence in this case to justify deviating from the default specifications. The product yields used in this analysis (Eq.~\ref{eq:wood_comb}) are for well-ventilated burning conditions. The simple chemistry reaction mechanism in FDS assumes constant, user-specified yields for carbon monoxide and soot. When a fire becomes underventilated, the yields of soot, carbon monoxide, or other product species change due to incomplete combustion. If the analysis of model results requires species concentrations, different species yields, a more complex chemical reaction, or finite-rate chemistry may be needed~\cite{FDS_Math_Guide}. The focus of this analysis is to examine the the effect on the flow conditions as a result of a interior door failure. Therefore, errors in species concentrations are tolerated.

To estimate the heat release rate per unit area (HRRPUA) of wood, Babrauskas and Grayson~\cite{babrauskas1990} conducted experiments in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials~\cite{ASTM:E1355}.} to determine the 5-min average of the HRRPUA for several different types of wood over a range of radiant heat fluxes. The results of that study indicate that the HRRPUA for different types of wood is approximately 50~kW/m$^2$. This value may be a conservative estimate for the HRRPUA for wood. As a result, the actual fire may have been larger. However, as discussed in this study, the fire used in the model produced interior conditions that were hazardous to fully protected firefighters.

@@

\subsection{Materials}
\label{matl}
While the fires were represented as fuel sources with a constant area (Section~\ref{fire}), from a heat transfer perspective, it is important to define the material properties (density, thermal conductivity, and specific heat) to account for heat transfer and energy storage in the ceiling, walls, and floor. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings on the main two floors of the structure, and the material properties of wood~\cite{Incropera:1} were specified on the unfinished portions of the attic and rear porch. Note that the multi-component heat transfer that would typically occur in an exterior wall (gypsum, wood studs, insulation, exterior finish, air gaps) was simplified and represented by a single material. The following lines define these materials in the FDS input file:

@@

In Fig.~\ref{fig:hrr}, the solid line represents the baseline FDS simulation (this follows the noted door openings from the incident). The dashed line represents an alternative series of events in which the rear door does not get opened and the subsequent breach of the rear porch wall on the left side of the structure does not occur. The dash-dotted line represents the prescribed (or nominal) HRR that was input into FDS. As shown in Fig.~\ref{fig:hrr}, the HRR output from FDS is less than the prescribed HRR because the fire in the structure becomes ventilation limited. The maximum HRR value is slightly less than 9~MW before the fire becomes ventilation limited and begins to decay. The HRR is reduced to 5~MW as the oxygen inside the structure is consumed. There is an insufficient amount of oxygen in the structure to burn all of the fuel, therefore combustion occurs locally at the vents where fresh air is available. Figure~\ref{fig:smv_ext_fire} shows flaming combustion through the failed window on the rear side of the structure and attic breach just after the second fire ignites at 40~s. Figure~\ref{fig:smv_ext_fire} also shows a photograph taken during an attic fire experiment that confirms this type of fire behavior. As a fire becomes underventilated, the majority of the combustion internal to the structure occurs on the rear porch where there is sufficient oxygen. Despite the lack of oxygen in the attic, the wood would continue to pyrolyze due to the high heat fluxes, therefore continue to produce additional gas phase fuel \cite{SFPE:Heskestad}. In the attic, combustion occurred at the external vents as a lack of oxygen present prevented combustion.  After additional wall breaches occur at 45~s and 55~s (see Table~\ref{tab:vents}), the abrupt change in ventilation causes the HRR to increase to 9~MW before it decreases to approximately 6~MW. The HRR then increases towards a sustainable steady-state value.

Figure~\ref{fig:hrr} indicates that the HRRs for both of the ventilation scenarios in FDS are identical until 130~s (HRR $\approx$ 7~MW). At 130~s, the rear door is opened in the baseline simulation. At 135~s, there is an additional breach on the left side of the structure (Table~\ref{tab:vents}). The open door and breach act as additional pressure releases for the structure. Again, the abrupt change in ventilation causes the HRR to increase from approximately 7~MW to 9.7~MW at 139~s. At the time of the door failure (160~s), the HRR in both cases was approximately 9~MW because of the increased amount of ventilation. While there might be some locally different burning characteristics between the baseline simulation and the alternative simulation, with respect to exposure to firefighters operating within flow path, both cases presented similar high hazard conditions. Therefore, the analysis will focus on the results of the baseline simulation.

\begin{figure}[!ht]
\centering
\includegraphics[width=.5\textwidth]{../Figures/smv_exterior_fire}
\includegraphics[width=.45\textwidth]{../Figures/PA020340}
\caption{Smokeview rendering of exterior combustion occurring on the rear side of the structure (left) and photograph from ongoing attic fire experiments (right).}
\label{fig:smv_ext_fire}
\end{figure}

\subsection{Pressure}
\label{pres}
As the simulated fire grows in the attic and rear porch, the pressure rises. Figure~\ref{fig:pres_159s} shows the calculated pressure conditions in the structure just before the interior door failed in the simulation. On the rear porch, the pressure rise ranges from a 4~Pa ($5.8 \times 10^{-4}$~psi) over-pressure at a height of 1~m to greater than 10~Pa ($1.5 \times 10^{-3}$~psi) at the ceiling from buoyancy due to temperature differences. Gases at an elevated pressure will always move towards a region of lower pressure.The door serves as a boundary between the high and low pressure regions. Once the door fails open, the high pressure gases flow towards the low pressure regions in the house, hallway, kitchen, and front stairwell. The velocity vectors shown in Fig.~\ref{fig:velo_175s} (15~s after the interior door fails) indicate the direction of flow within the structure.
